\documentclass{amsart}

\usepackage{amsmath}
\usepackage{tensor}

\DeclareMathOperator{\cf}{cf}

\title{On a class of Polish-like spaces}

\author{Claudio Agostini}
\author{Luca Motto Ros}

\address[Claudio Agostini]{Universit\`a degli Studi di Torino, Dipartimento di Matematica ``G. Peano'', Via Carlo Alberto 10, 10123 Torino, Italy.}
\email[Claudio Agostini]{claudio.agostini@unito.it}

\address[Luca Motto Ros]{Universit\`a degli Studi di Torino, Dipartimento di Matematica ``G. Peano'', Via Carlo Alberto 10, 10123 Torino, Italy.}
\email[Luca Motto Ros]{luca.mottoros@unito.it}

%\curraddr{curraddr}
%\email{email}
%\urladdr{urladdr}
\thanks{The work was supported in part by 
Universit\`a degli Studi of Torino (Italy), in part by the Gruppo Nazionale per le Strutture Algebriche, Geometriche e le loro Applicazioni (GNSAGA) of the Istituto Nazionale di Alta Matematica (INdAM) of Italy.}
\keywords{Generalized descriptive set theory, general topology, topological games, strong Choquet spaces.}
\begin{document}

\maketitle


The Cantor and Baire spaces have natural generalizations to uncountable cardinals $\kappa$, and in the last two decades many results have been proven about them under the hypothesis $\kappa^{<\kappa}=\kappa$. There is more uncertainty about what should be the right generalization of Polish spaces. What we expect is a class of spaces of weight $\kappa$ which includes the two previous mentioned and can support most of generalized descriptive set theory.
Recently S. Coskey and P. Schilcht in \cite{CosSch} showed that one property that characterize the completeness of metric spaces in the classical case, being strong Choquet, can be extended to higher cardinals and gives a class of spaces with promising properties. 


The assumption $\kappa^{<\kappa}=\kappa$ is equivalent to $2^{<\kappa}=\kappa$ for $\kappa$ regular, but the second one is suitable also for singular cardinals. 
In a forthcoming paper, V. Dimonte, L. Motto Ros and X. Shi use this last assumption to study descriptive set theory on singular cardinals of countable cofinality, where there is a natural definition of $\lambda$-Polish as a completely metrizable space of weight $\lambda$. 

In this talk, I will present an ongoing work in collaboration with L. Motto Ros where we try to put together this two approaches and study a class of spaces which is suitable for all cardinals satisfying $2^{<\lambda}=\lambda$, but coincide with S. Coskey and P. Schlicht's definition if $\lambda$ is regular, and with V. Dimonte, L. Motto Ros and X. Shi's one if $\lambda$ has countable cofinality. I will define this class and show how many properties that holds in the classical theory may be proved with different or similar tools in this case as well.  

\begin{thebibliography} {9}

\bibitem{CosSch} S. Coskey and P. Schlicht, \emph{Generalized Choquet spaces}, Fund. Math., \textbf{232} (2016), 227--248.

\bibitem{Kechris} A. Kechris, \emph{Classical descriptive set theory}, Springer Science \& Business Media, 2012.
 
\bibitem{Engelking} R. Engelking, \emph{General Topology}, Heldermann Verlag, Berlin, second ed., 1989.
 
\end{thebibliography}



\end{document}