
\documentclass[10pt]{amsart}
\usepackage{amsmath,amssymb,amscd,amsthm,amsfonts,amstext,amsbsy,mathrsfs,hyperref,upgreek,mathtools,stmaryrd,enumitem,bbm, mathabx}
%\usepackage{MnSymbol}
%\usepackage[shadow]{todonotes}
% \usepackage{citeref, breakcites}

\usepackage[
%disable, 
textsize=footnotesize,color=yellow!40, bordercolor=white]{todonotes}

\usepackage%[scale=0.682]
[a4paper, margin=1.2in]{geometry}

\usepackage{xcolor}
\definecolor{blue}{rgb}{0,0,1}
\hypersetup{
	unicode=true,
	colorlinks=true,
	citecolor=blue,
	linkcolor=blue,
	anchorcolor=blue
}

\hypersetup{colorlinks=true}

% other fonts
\newcommand\V{\mathsf{V}}
\newcommand{\Minimal}{\MM[\![G]\!]}
\newcommand\bB{\boldsymbol{{\rm B}}}
\newcommand\bN{\boldsymbol{{\rm N}}}
\renewcommand{\L}{\mathcal{L}}
%\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\EE}{\mathbb{E}}                                                       
\newcommand{\OO}{\mathbb{O}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\HH}{\mathbb{H}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\Ss}{\mathbb{S}}
\newcommand{\HHH}[1]{\mathrm{H}({#1})}
\newcommand{\Hes}{\mathfrak{H}}
\newcommand{\pre}[2]{{}^{#1} #2}
\newcommand{\seq}[2]{\langle #1 \mid #2 \rangle}
\newcommand{\set}[2]{\{ #1 \mid #2 \}}
\newcommand{\Nbhd}{\mathbf{N}}
\newcommand{\Bor}{\mathsf{Bor}}
\newcommand{\pI}{\mathrm{I}}
\newcommand{\pII}{\mathrm{II}}
%\newcommand{\K}[1]{\boldsymbol{K}_{#1}}
\newcommand{\anf}[1]{{\text{``}\hspace{0.3ex}{#1}\hspace{0.01ex}\text{''}}}
\newcommand{\op}{\mathsf{op}}
\newcommand{\rnk}{\mathsf{rnk}}

% operators
\newcommand{\id}{\operatorname{id}}
\newcommand{\cof}{\operatorname{cof}}
\newcommand{\otp}[1]{{{\rm{otp}}\left(#1\right)}}
\newcommand{\supp}[1]{{{\rm{supp}}(#1)}}
\newcommand{\proj}{\operatorname{p}}
\newcommand{\pow}{\mathscr{P}}
\newcommand{\Trcl}{\operatorname{Trcl}}
\newcommand{\p}{\operatorname{p}}
\newcommand{\leng}{\operatorname{lh}}
\newcommand{\range}{\operatorname{range}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\Col}{\operatorname{Col}}
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\Llr}{\Longleftrightarrow}
\newcommand{\tc}{\operatorname{tc}}
\newcommand{\depth}{\operatorname{d}}
%\newcommand{\Add}{\operatorname{Add}}
\newcommand{\ul}{\ulcorner}
\newcommand{\ur}{\urcorner}
\newcommand{\lh}{\operatorname{lh}}
\newcommand{\llangle}{\langle\!\langle}
\newcommand{\rrangle}{\rangle\!\rangle}
\newcommand{\vn}{\vec{n}}
\newcommand{\one}{\mathbbm1}
\newcommand{\lkk}{\L_{\kappa,\kappa}}
\newcommand{\gbr}[1]{{\ulcorner{#1}\urcorner}}
\newcommand{\Def}{\mathrm{Def}}
% spaces
\newcommand{\On}{{\mathrm{Ord}}}
%\newcommand{\Card}{{\sf Card}}
\newcommand{\Lim}{{\mathrm{Lim}}}
\newcommand{\Fm}{{\mathrm{Fml}}}
%\newcommand{\Sym}{{\rm Sym}}
\newcommand{\push}{{\mathrm{push}}}

\newcommand{\spl}{\mathrm{split}}


% axioms
\newcommand{\AD}{{\sf AD}}
\newcommand{\ZFC}{{\sf ZFC}}
\newcommand{\ZF}{{\sf ZF}}
\newcommand{\AC}{{\sf AC}}
\newcommand{\DC}{{\sf DC}}
\newcommand{\KM}{{\sf KM}}
\newcommand{\GB}{{\sf GB}}
\newcommand{\GBC}{{\sf GBC}}

\newcommand{\ran}{\mathrm{ran}} 

% others
\newcommand{\Linf}{\L_{\omega_1 \omega}}

%Luecke
\newcommand{\map}[3]{{#1}:{#2}\longrightarrow{#3}}
\newcommand{\Map}[5]{{#1}:{#2}\longrightarrow{#3};~{#4}\longmapsto{#5}}
\newcommand{\pmap}[4]{{#1}:{#2}\xrightarrow{#4}{#3}}
\newcommand{\Set}[2]{\{{#1}~\vert~{#2}\}}
\newcommand{\dom}[1]{{{\rm{dom}}(#1)}}
\newcommand{\length}[1]{{{\rm{lh}}(#1)}}
\newcommand{\VV}{{\rm{V}}}
\newcommand{\WW}{{\rm{W}}}
\newcommand{\LL}{{\rm{L}}}
\newcommand{\Add}[2]{{\rm{Add}}({#1},{#2})}
\newcommand{\goedel}[2]{{\prec}{#1},{#2}{\succ}}
\newcommand{\ot}{ot}




\newcommand{\ZZ}{\mathbb{Z}} 
\newcommand{\QQ}{\mathbb{Q}} 
\newcommand{\RR}{\mathbb{R}} 
\newcommand{\GG}{\mathbb{G}}
\newcommand{\CCC}{\mathbb{C}} 
\newcommand{\NN}{\mathbb{N}} 

\newcommand{\NNN}{\mathcal{N}} 
\newcommand{\MM}{\mathcal{M}}
\newcommand{\KK}{\mathcal{K}} 
\renewcommand{\AA}{\mathcal{A}}
\newcommand{\BB}{\mathcal{B}}
\newcommand{\CC}{\mathcal{C}} 
\newcommand{\DD}{\mathcal{D}} 

%\newcommand{\MM}{\mathbb{M}}
%\newcommand{\NN}{\mathbb{N}} 
%\newcommand{\KK}{\mathbb{K}} 
%\renewcommand{\AA}{\mathbb{A}}
%\newcommand{\BB}{\mathbb{B}}
%\newcommand{\CC}{\mathbb{C}} 
%\newcommand{\DD}{\mathbb{D}} 




\newcommand{\K}{\mathcal{K}} 
\newcommand{\Th}{\mathrm{Th}} 

\newcommand{\DLO}{\mathrm{DLO}} 
\newcommand{\ACF}{\mathrm{ACF}} 

\newcommand{\lex}{\mathrm{lex}} 

\newcommand{\cC}{\mathcal{C}} 
\newcommand{\Mod}{\mathrm{Mod}} 

\newcommand{\acl}{\mathrm{acl}} 
\newcommand{\tp}{\mathrm{tp}} 
\newcommand{\Aut}{\mathrm{Aut}} 
\newcommand{\Sym}{\mathrm{Sym}} 

\newcommand{\Lev}{\mathrm{Lev}} 



%\setcounter{secnumdepth}{2}
% theorems
\newtheorem{theorem}{Theorem}[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{claim}[theorem]{Claim}
\newtheorem*{claim*}{Claim}
\newtheorem*{subclaim*}{Subclaim}
\newtheorem*{Vaughtconj}{Vaught's Conjecture}
\theoremstyle{definition}
%\newtheorem{claim}{Claim}[theorem]
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{definition*}{Definition}
\newtheorem*{notation}{Notation}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{example}[theorem]{Example}
\newtheorem*{example*}{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}



% enumeration

\newenvironment{enumerate-(a)}{\begin{enumerate}[label={\upshape (\alph*)}, leftmargin=2pc]}{\end{enumerate}}

\newenvironment{enumerate-(a)-r}{\begin{enumerate}[label={\upshape (\alph*)}, leftmargin=2pc,resume]}{\end{enumerate}}

\newenvironment{enumerate-(A)}{\begin{enumerate}[label={\upshape (\Alph*)}, leftmargin=2pc]}{\end{enumerate}}

\newenvironment{enumerate-(A)-r}{\begin{enumerate}[label={\upshape (\Alph*)}, leftmargin=2pc,resume]}{\end{enumerate}}

\newenvironment{enumerate-(i)}{\begin{enumerate}[label={\upshape (\roman*)}, leftmargin=2pc]}{\end{enumerate}}

\newenvironment{enumerate-(i)-r}{\begin{enumerate}[label={\upshape (\roman*)}, leftmargin=2pc,resume]}{\end{enumerate}}

\newenvironment{enumerate-(I)}{\begin{enumerate}[label={\upshape (\Roman*)}, leftmargin=2pc]}{\end{enumerate}}

\newenvironment{enumerate-(I)-r}{\begin{enumerate}[label={\upshape (\Roman*)}, leftmargin=2pc,resume]}{\end{enumerate}}

\newenvironment{enumerate-(1)}{\begin{enumerate}[label={\upshape (\arabic*)}, leftmargin=2pc]}{\end{enumerate}}

\newenvironment{enumerate-(1)-r}{\begin{enumerate}[label={\upshape (\arabic*)}, leftmargin=2pc,resume]}{\end{enumerate}}

\newenvironment{itemizenew}{\begin{itemize}[leftmargin=2pc]}{\end{itemize}}




\begin{document}

%\thanks{The first and third author were partially supported by DFG-grant LU2020/1-1.}

%\subjclass[2010]{03E30, 03E40, 03E70} 

%\keywords{Class forcing, Separation, Replacement} 


\author{Philipp Schlicht}
\address{Philipp Schlicht, School of Mathematics, University of Bristol, Fry Building, Woodland Road, Bristol, BS8 1UG, UK}
\email{philipp.schlicht@bristol.ac.uk}
\urladdr{}

%\title{}
\date{\today}


\title[Basics of descriptive set theory]{The very basics of descriptive set theory}

%\begin{abstract} t
%Lecture notes from the lecture Advanced mathematical logic: Model theory, Bonn October-November 2017. 
%\end{abstract} 

\maketitle


\setcounter{tocdepth}{2}
%\tableofcontents 


%\newpage 



%%%%%%%%%%%%
%%%%%%%%%%%%
%\section{Basics of descriptive set theory} 

\begin{abstract} 
This note presents elementary proofs that any two uncountable Polish metric spaces are Borel isomorphic, a function is Borel measurable if and only if its graph is a Borel set, and the auxiliary result that disjoint analytic sets can be separated by Borel sets. 
The proofs are much shorter than those in the literature. 
%The proofs are elementary, but stand in contrast with those in the literature, which usually re
%much shorter than proofs in the literature and are obtained by removing all redundancies. 
%The proofs are elementary, but stand in contrast with those in the literature, which usually refer to a number of other results. 
\end{abstract} 

%In section, we review some material on standard Borel spaces, Borel sets, analytic sets and Borel measurable functions, in particular the fact that any two uncountable standard Borel spaces are isomorphic. 
%We give proofs from scratch. 
%We give very direct and concise proofs from scratch, in contrast to most proofs in the literature, which refer to other theorems. 

%%%%%%%%%%%%
\section{Standard Borel spaces} 

\begin{definition} 
\begin{enumerate-(a)} 
\item 
A metric space $(X,d)$ is called \emph{Polish} if it is countably based and complete. 
\item 
A topological space $(X,\tau)$ is called \emph{Polish} if there is a Polish metric $d$ on $X$ that induces $\tau$. 
\item 
A pair $(X,\sigma)$ is called a \emph{standard Borel space} if $\sigma$ is the Borel $\sigma$-algebra generated by a Polish topology on $X$.  
\end{enumerate-(a)} 
\end{definition} 

Our notation is $(X,d)$, $(Y,d)$, $(Z,d)$ for Polish metric spaces and the metric $d$ is often omitted. 
We further write $d(A)$ for the \emph{diameter} $\sup_{x,y\in A} d(x,y)$ of a subset $A$ of $X$, and $B_\epsilon(x)$ for the open ball of radius $\epsilon$ around $x$. 

\begin{example} 
$\RR^n$, Baire space $\NN^\NN$,  Cantor space $2^\NN$, compact metric spaces, countably based locally compact metric spaces. 
\end{example} 

Here $\NN^\NN$ caries the product topology. 
The standard metric on $\NN^\NN$ is defined by $d(x,y)=\frac{1}{2^n}$ for the least $n$ with $x(n)\neq y(n)$, where $x\neq y$. 
The basic open subsets of $\NN^\NN$ (clopen balls) are denoted $N_t=\{x\in \NN^\NN\mid t\subseteq x\}$ for $t\in \NN^{<\NN}$. 
The following combinatorial presentation of $\NN^\NN$ via trees is useful for constructions of continuous maps. 
A \emph{tree} on $\NN$ is a subset of $\NN^{<\NN}$ that is closed under initial segments. 
It is easy to see that the closed subsets of $\NN^\NN$ are exactly the sets of the form $[T]=\{x\in \NN^\NN\mid \forall n\ x{\upharpoonright}n \in T\}$ for trees $T$ on $\NN$. 
Let $|t|$ denote the length of a sequence $t\in \NN^{<\NN}$. 
Let $x{\upharpoonright}n$ denote the restriction of a sequence $x\in \NN^\NN$ to $\{0,\dots,n-1\}$. 
A tree is called \emph{pruned} if it has no end nodes. 
Let $\Lev_n(T)$ denote the $n$-th level of a tree $T$, so the elements of $\Lev_n(T)$ are sequences of length $n$. 
%where $\Lev_0(T)$ consists only of the root. 
%A node $t\in T$ \emph{splits} in $T$ if it has at least $2$ direct successors in $T$. Let $\spl(T)$ denote the set of splitting nodes of $T$. A tree $T$ is called \emph{perfect} if it is nonempty, pruned and every node in $T$ has some node in $\spl(T)$ above it. 

\bigskip 
Note that Polish spaces are closed under countable products. 
To see this, take  a sequence $\vec{X}=\langle X_n\mid n\in\NN\rangle$ of Polish spaces and assume that each metric $d$ is bounded by replacing $d$ with $\max\{d,1\}$. 
Then form the weighted sum over all coordinates with factors $2^{-n}$. 
It is easy to show that this metric on the product is complete and induces the product topology. 

\bigskip 
We next show that any two uncountable Polish spaces are \emph{Borel isomorphic} in the sense of a bijection that preserves Borel sets both ways. 

%\begin{definition} 
%A \emph{Borel isomorphism} $f\colon A\rightarrow B$ between Borel subsets $A$ and $B$ of standard Borel spaces is a bijection that preserves Borel sets in both directions. 
%\end{definition} 

\begin{proposition} \label{Borel isomorphism of Polish spaces} 
Any two uncountable Polish spaces are Borel isomorphic. 
\end{proposition} 

This will follow from the next two lemmas. 
To state them, we fix some notation. 
%The first three lemmas show that any uncountable Polish space $X$ is Borel isomorphic to a Borel subset of $\NN^\NN$ and vice versa. The last lemma show that this is sufficient to obtain Proposition \ref{Borel isomorphism of Polish spaces}. 
%\begin{definition} 
%\begin{enumerate-(a)} 
%\item 
An \emph{$\epsilon$-cover} of a subset $A$ of $X$ is a sequence $\vec{A}=\langle A_i\mid i<N\rangle$ with $N\in \NN\cup\{\NN\}$ whose union equals $A$ with $d(A_i)<\epsilon$ and $\overline{A_i}\subseteq A$ for all $i\in\NN$. 
%\item 
%Let $T$ be a pruned subtree of $\NN^{<\NN}$ and $\vec{\epsilon}=\langle \epsilon_i \mid i\in\NN\rangle$ be any sequence in $\RR^+$ converging to $0$. 
%A \emph{$T$-scheme} for $X$ is a family $\vec{A}=\langle A_t\mid t\in T\rangle$ of nonempty subsets of $X$ with $A_0=X$ that satisfies the following properties for all $t\in T$: \\ 
% (i) $A_t=\bigcup_{ti \in T} A_{ti}$ (disjoint), \ \ (ii) $\overline{A_{ti}} \subseteq A_t$ if $ti\in T$, and \ \ (iii) $d(A_t)<\epsilon_n$ if $|t|=n$. 
%\end{enumerate-(a)} 
%\end{definition} 
An \emph{$F_\sigma$-set} is a countable union of closed sets. 
Note that all open sets are $F_\sigma$. 

\begin{lemma} \label{partition epsilon-covers} 
Let $\epsilon>0$. 
\begin{enumerate-(1)} 
\item \label{partition epsilon-covers - open sets} 
Any open set has an $\epsilon$-cover consisting of open sets. 
\item 
Any $F_\sigma$-set has a disjoint $\epsilon$-cover consisting of $F_\sigma$-sets. 
\end{enumerate-(1)} 
\end{lemma} 
\begin{proof} 
(1) 
Suppose that $A$ is an open subset of $X$. 
Fix a countable dense subset $Q$ of $X$ and let $\vec{A}=\langle A_i\mid i\in N\rangle$ list all open balls $B_q(x)$ with $x\in Q$, $q\in (0,\frac{\epsilon}{2})\cap \QQ$ and $\overline{B_q(x)}\subseteq A$. 
It suffices to show that $A=\bigcup_{i\in N} A_i$. To see this, take any $x\in A$ and find some $q\in \QQ$ with $B_{2q}(x)\subseteq A$. 
Since $Q$ is dense, there is some $y\in B_q(x)\cap Q$. We then also have $x\in B_q(y)$ and thus it suffices to show that $\overline{B_q(y)}\subseteq A$. 
This holds since $\overline{B_q(y)}\subseteq B_{2q}(x)\subseteq A$. 
%Then $x\in B_q(y)$ and $\overline{B_q(y)}\subseteq A$. 
%By the latter, $B_q(y)$ appears in $\vec{A}$. 

(2) 
Suppose that $A$ is an $F_\sigma$ subset of $X$. 
Take a sequence $\vec{A}=\langle A_i\mid i\in N\rangle$ whose union equals $A$ with $A_i$ closed. 
Then $A$ is the disjoint union of the sets $B_i=A_i\setminus (A_0\cup\dots\cup A_{i-1})$. 
Since $B_i$ is of the form $C\cap D$ with $C$ open and $D$ closed, it suffices to prove the claim for sets of this form. 
To see this, note that by (a), there is an $\epsilon$-cover $\vec{C}=\langle C_i\mid i\in K\rangle$ of $C$ consisting of open sets. 
Let $\vec{D}=\langle D_i\mid i\in K\rangle$ with $D_i=(C_i\setminus (C_0\cup\dots\cup C_{i-1}))\cap D$. 
Then $\vec{D}$ is a disjoint $\epsilon$-cover of $C\cap D$ consisting of $F_\sigma$-sets. 
\end{proof} 

%The set $X^\sqbullet$ of $x\in X$ such that all neighborhoods of $x$ are uncountable is called the \emph{perfect kernel} of $X$. 
%Since $X^\sqbullet$ is obtained by removing countably many basic open sets, each of which contains only countably many elements of $X$, it is a Borel subset of $X$. 

%The next lemma yields a Borel isomorphism between $X$ and $\NN^\NN$ by working with $X^\sqbullet$ instead of $X$ and noting that countable sets do not matter for the existence of Borel isomorphisms. 
%For the proof of this fact, we only need the second part of the lemma, but the first part will be useful later. 

The existence of Borel isomorphisms between uncountable Polish spaces follows from the second part of the next lemma by noting that countable sets can be discarded. 
We call a subset of $X$ \emph{co-countable} if its complement is countable. 

\begin{lemma} \label{Polish space as continuous image} \ 
%Any Polish space $X$ is Borel isomorphic to a Borel subset of $\NN^\NN$. 
\begin{enumerate-(1)} 
\item 
There is a closed subset $A$ of $\NN^\NN$ and a continuous Borel isomorphism $f\colon A\rightarrow X$.\footnote{In fact, any continuous bijection is a Borel isomorphism, but we don't prove this here. } 
\item 
If $X$ is uncountable, then there is a continuous Borel isomorphism $f\colon \NN^\NN\rightarrow X$ onto a co-countable subset of $X$. 
\end{enumerate-(1)} 
\end{lemma} 
\begin{proof} 
(1) 
Let $\vec{\epsilon}=\langle \epsilon_i \mid i\in\NN\rangle$ be a sequence in $\RR^+$ converging to $0$. 
By iterative applications of Lemma \ref{partition epsilon-covers}, we obtain a tree $T\subseteq \NN^{<\NN}$ without end nodes and a family $\vec{A}=\langle A_t\mid t\in T\rangle$ of nonempty $F_\sigma$-subsets of $X$ with $A_0=X$ that satisfies the following properties for all $t\in T$: 
%\\ 
%\indent 
(a) $A_t=\bigcup_{ti \in T} A_{ti}$ is a disjoint union, \ \ (b) $\overline{A_{ti}} \subseteq A_t$ for $ti\in T$, and \ \ (c) $d(A_t)<\epsilon_n$ if $|t|=n$. 
%\\ 
%\begin{enumerate-(a)} 
%\item 
%$A_t=\bigcup_{ti \in T} A_{ti}$ (disjoint), 
%\item 
%$\overline{A_{ti}} \subseteq A_t$ and 
%\item 
%$d(A_t)<\epsilon_n$ if $|t|=n$. 
%\end{enumerate-(a)} 
%Note that $A:=[T]$ is a closed subset of $\NN^{<\NN}$. 

We then define $f\colon [T]\rightarrow X$ by choosing $f(x)$ as the unique element of $\bigcap_{n\in\NN} A_{x\upharpoonright n}$. 
The latter is nonempty by (b), (c) and completeness of $X$, and has a unique element by (c). 

\begin{claim*} 
$f$ is bijective. 
\end{claim*} 
\begin{proof} 
It follows from (a) that $f$ is injective. 
To see that $f$ is surjective, take any $x\in X$. 
By (a), there is for each $n\in\NN$ a unique $t_n\in T$ of length $n$ with $x\in A_{t_n}$. 
By (b), $t_i\subseteq t_j$ for $i\leq j$. 
Hence $f(y)=x$ for $y=\bigcup_{i\in\NN} t_i$. 
\end{proof} 

\begin{claim*} 
$f$ is continuous. 
\end{claim*} 
\begin{proof} 
If $U$ is open and $f(x)\in U$, then $A_{x{\upharpoonright}n}\subseteq U$ for sufficiently large $n$ by (c). 
\end{proof} 

\begin{claim*} 
$f$ is a Borel isomorphism. 
\end{claim*} 
\begin{proof} 
Since $f$ is continuous, it suffices that the images of Borel sets are Borel. 
This holds since $f$ maps each basic open set $N_t$ onto $A_t$ and $f$ is bijective. 
\end{proof} 
Thus $f$ is as required. 

\bigskip 

(2) 
A slight modification of the construction in (a) works. 
We will ensure that $A_t$ is uncountable for all $t\in T$. 
The change is that we remove a set $A_{ti}$ if it is countable. Note that one can easily ensure that $A_{ti}$ is uncountable for infinitely many $i\in\NN$ by partitioning the sets further. Therefore one can assume that $ti\in T$ for all $i\in\NN$ and thus $T=\NN^{{<}\NN}$. 
During the construction only a countable set is removed in total, yielding the claim. 
\end{proof} 

Note that in (2), for any $X$ with isolated points, one cannot replace the co-countable set by $X$. 


\iffalse 
Using the previous lemma, one can see that any nonempty Polish space $X$ is the image of $\NN^\NN$ under a continuous function. 
By the lemma, it suffices to show this in the case that $X=[T]$ is a closed subset of $\NN^\NN$. In this case, map each $x\in \NN^\NN$ to the leftmost $y\in [T]$ with $x{\upharpoonright}n \subseteq y$, where $n$ is maximal with $x{\upharpoonright}n\in T$. 

\begin{lemma} \todo{This and the next lemma can then be omitted } 
If $X$ is an uncountable Polish space, then $\NN^\NN$ is Borel isomorphic to a Borel subset of $X$. 
\end{lemma} 
\begin{proof} 
Fix a sequence $\vec{\epsilon}=\langle \epsilon_i \mid i\in\NN\rangle$ in $\RR^+$ converging to $0$. 
It is easy to construct a family $\vec{U}=\langle U_t\mid t\in \NN^{<\NN}\rangle$ of uncountable open subsets of $X$ such that: 
(a) $\overline{U_t}\subseteq U_s$ for all $s\subseteq t$, \\ 
(b) $U_s\cap U_t=\emptyset$ for all $s\neq t$ with $|s|=|t|$, and \ \  (c) $d(U_t)<\epsilon_n$ if $|t|=n$. 
%$\lim_{i\rightarrow \infty} d(U_{ti})=0$ for all $t\in \NN^{<\NN}$. 
% and $\lim_{i\rightarrow \infty} x_i\in Q$ for any sequence $\vec{x}=\langle x_i\mid i\in\NN\rangle$ with $x_i\in U_{ti}$ for all $i\in\NN$. 

Define $f\colon \NN^\NN\rightarrow X$ by choosing $f(x)$ as the unique element of $\bigcap_{n\in \NN} U_{x{\upharpoonright}n}$. 
It is then easy to see that $f$ is a homeomorphism onto its image. 
To see that $f(\NN^\NN)$ is Borel, let $U_n=\bigcup_{|t|=n} U_t$. We then have $f(\NN^\NN)=\bigcap_{n\in\NN} U_n$. 
\end{proof} 

\begin{lemma} 
Suppose that $X$ is Borel isomorphic to a Borel subset of $Y$ and vice versa. 
Then $X$ and $Y$ are Borel isomorphic. 
\end{lemma} 
\begin{proof} 
Suppose that $f\colon X\rightarrow Y$ and $g\colon Y\rightarrow X$ are Borel isomorphisms onto Borel sets. 
We call a pair $(A,B)$ of Borel subsets of $X$ and $Y$ \emph{nice} if $f\colon A\rightarrow B$ and $g\colon B\rightarrow A$ are well-defined.\footnote{This is shorthand for $\ran(f{\upharpoonright}A)\subseteq B$ and $\ran(g{\upharpoonright}B)\subseteq A$. } 

\begin{claim*} 
For any nice pair $(A,B)$, there is a nice pair $(A_*,B_*)$ such that $A_*\subseteq A\cap g(B)$, $B_*\subseteq B\cap f(A)$ and the Borel sets $A\setminus A_*$, $B\setminus B_*$ are Borel isomorphic. 
\end{claim*} 
\begin{proof} 
To see this, let $A^*=A\setminus g(B)$, $B^*=B\setminus f(A)$. 
Since $A^*\cap g(B^*)=B^*\cap f(A^*)=\emptyset$, the maps $f\colon A^*\rightarrow f(A^*)$ and $g\colon B^*\rightarrow g(B^*)$ witness that 
$A^*\cup g(B^*)$ and $B^*\cup f(A^*)$ are Borel isomorphic. 
Thus $A_*=A\setminus (A^*\cup g(B^*))=g(B) \setminus g(B^*)$ and $B_*=B\setminus (B^*\cup f(A^*))=f(A)\setminus f(A^*)$ are as required. 
\end{proof} 

Let $A_0=X$ and $B_0=Y$. By iterative applications of the previous claim, we obtain decreasing sequences $\vec{A}=\langle A_n\mid n\in\NN\rangle$ and $\vec{B}=\langle B_n\mid n\in\NN\rangle$ of Borel sets such that $(A_n, B_n)$ is nice for all $n\in\NN$. Let $A=\bigcap_{n\in\NN} A_n$ and $B=\bigcap_{n\in\NN}B_n$. 

It suffices to show that $A$ and $B$ are Borel isomorphic via $f$. 
To see this, note that $f\colon A\rightarrow B$ is well-defined, since $f\colon A_n\rightarrow B_n$ is well-defined for all $n\in\NN$. 
Moreover, $f\colon A\rightarrow B$ is surjective, since $B_{n+1}\subseteq f(A_n)$ for all $n\in \NN$. 
\end{proof} 

\fi 





%%%%%%%%%%%%
\section{Analytic sets} 

Let $p\colon X\times Y\rightarrow X$ denote the first projection. 

\begin{definition} 
A subset $A$ of $X$ is called \emph{analytic} if $A=p(B)$ for some Borel subset $B$ of $X\times Y$ for some standard Borel space $Y$. 
\end{definition} 

Clearly this class contains all Borel sets. 

By Proposition \ref{Borel isomorphism of Polish spaces}, one can take $Y$ to be a fixed uncountable Polish space. 
Moreover, the analytic subsets of an uncountable standard Borel space $X$ are precisely the images of analytic subsets of $\NN^\NN$ under Borel isomorphisms. 
Hence we can work in $\NN^\NN$ to study their properties. 
To do this, we will go through a slight detour. 
We call a subset $A$ of $\NN^\NN$ \emph{${}^*$analytic} if $A=p[C]$ for some closed subset $C$ of $\NN^\NN\times \NN^\NN$, equivalently $A=p[T]=\{x\mid \exists y\ (x,y)\in [T]\}$ for some subtree $T$ of $(\NN\times\NN)^{<\NN}$. 

\begin{lemma} \label{characterization of analytic sets} 
\begin{enumerate-(1)} 
\item \label{characterization of analytic sets - closure} 
The class of ${}^*$analytic sets is closed under countable unions and intersections. 
%It follows that every Borel subset of $\NN^\NN$ is ${}^*$analytic. 
%\emph{(Hint: first show that the intersection of two ${}^*$analytic sets is again ${}^*$analytic by merging trees projecting to the sets.) } 
\item 
Every analytic subset of $\NN^\NN$ is ${}^*$analytic. 
%\emph{(Hint: this is an easy consequence of (a).) } 
\end{enumerate-(1)} 
\end{lemma} 
\begin{proof} 
(1) 
Closure under countable unions is easily seen by forming unions of disjointified trees. 

Closure under countable intersections is proved by merging trees. 
We do only the case of two trees, since the general case is similar. 
So suppose that $T_0$ and $T_1$ are subtrees of $(\NN\times \NN)^{{<}\NN}$. 
We identify elements of $(\NN\times \NN)^{{<}\NN}$ with pairs of sequences. 
For $t\in \NN^{{<\NN}}$, define $t^0, t^1\in \NN^{{<}\NN}$ by letting $t^0(i)=t(2i)$ if $2i<k$ and $t^1(i)=t(2i+1)$ if $2i+1<k$. 
Let $U$ be the subtree of $(\NN\times \NN)^{{<}\NN}$ that consists of all $(s,t)\in (\NN\times \NN)^k$ with $(s,t^0)\in S$ and $(s,t^1)\in T_1$. 
One can now check that $p[S]\cap p[T]=p[U]$. 

(2) 
The collection of subsets generated from basic open sets by countable unions and intersections forms a $\sigma$-algebra and thus equals the Borel $\sigma$-algebra. 
Hence all Borel sets are ${}^*$analytic. 
The claim follows by composing two projections. 
\end{proof} 


The previous result also holds for finite products of $\NN^\NN$, by observing that the proof still works or noting that they are homeomorphic to $\NN^\NN$. 

In the next proof, subsets $A$, $B$ of $X$ are called \emph{Borel separated} if there is a Borel set $C$ with $A\subseteq C$ and $B\subseteq X\setminus C$. 

\begin{proposition} \label{separation of analytic sets} 
Any two disjoint analytic subsets $A$, $B$ of $X$ are Borel separated. 
\end{proposition} 
\begin{proof} 
It suffices to show this for $X=\NN^\NN$, and by Problem \ref{characterization of analytic sets} $A$ and $B$ are then *analytic. 
It follows that there are continuous surjections $f\colon [S]\rightarrow A$ and $g\colon [T]\rightarrow B$, where $S$ and $T$ are pruned subtrees of $\NN^{<\NN}$. 
%\footnote{One can assume that $[S]=[T]=\NN^\NN$ to simplify the notation at the expense of using Lemma \ref{analytic sets as continuous images of N^N}. } 
It is an easy to see that there is a continuous surjection from $\NN^\NN$ onto $[S]$, in fact one that fixes all elements of $[S]$. 
We can thus assume that $[S]=[T]=\NN^\NN$. 
%\todo{We can assume $[S]=[T]=\NN^\NN$ anyway, by using retractions from $\NN^\NN$ onto them. } 

Suppose that $A$, $B$ are not Borel separated. 
We then construct increasing sequences $\vec{s}=\langle s_i\mid i\in\NN\rangle$ and $\vec{t}=\langle t_i\mid i\in\NN\rangle$ in $\NN^{<\NN}$ such that for all $i\in\NN$, $f(N_{s_i})$ and $g(N_{t_i})$ are not Borel separated as follows. 

Let $s_0=t_0=\emptyset$. 
Given $s_i$ and $t_i$, there are $k,m\in \NN$ such that $f(N_{s_i k})$ and $g(N_{t_i m})$ are not Borel separated. 
If each of these pairs were Borel separated by some $B_{k,m}$, then $C_k=\bigcap_{t_i m\in T} B_{k,m}$ separates $f(N_{s_i k})$ and $g(N_{t_i})$ for each $k\in\NN$, and hence $C=\bigcup_{s_i k\in S} C_k$ separates $f(N_{s_i})$ and $g(N_{t_i})$, contradicting our assumption. 
Let $s_{i+1}=s_i k$ and $t_{i+1}= t_i m$. 

Let $x=\bigcup_{i\in\NN} s_i\in [S]$ and $y=\bigcup_{i\in\NN} t_i\in [T]$. 
Since $A$ and $B$ are disjoint, $f(x)\neq g(y)$. 
Let further $U$, $V$ be disjoint open subsets of $X$ with $f(x)\in U$ and $g(y)\in V$. 
By continuity of $f$ and $g$, $f(N_{x{\upharpoonright}n})\subseteq U$ and $g(N_{y{\upharpoonright}n})\subseteq V$ for sufficiently large $n\in\NN$, so $f(N_{x{\upharpoonright}n})$ and $g(N_{y{\upharpoonright}n})$ are Borel separated. 
However, they are not Borel separated by the construction. 
%But this contradicts the construction. 
\end{proof} 

Next is a very useful characterization of analytic sets. 

\begin{lemma} \label{analytic sets as continuous images of N^N} 
Every nonempty analytic subset of $X$ is the range of some continuous $f\colon \NN^\NN\rightarrow X$. 
\end{lemma} 
\begin{proof} 
%Since every nonempty Polish space is a continuous image of $\NN^\NN$ by the remark after Lemma \ref{Polish space as continuous image}, it suffices to find some continuous $f\colon Y\rightarrow X$ with $f(Y)=X$. 
Take an analytic subset $A$ of $X$. By Lemma \ref{Polish space as continuous image}, there are a closed subset $Y$ of $\NN^\NN$ and a continuous Borel isomorphism $f\colon Y\rightarrow X$. 
It suffices to find a continuous bijection $g\colon \NN^\NN\rightarrow f^{-1}(A)$. 
Then $A$ equals the range of $fg\colon \NN^\NN\rightarrow X$. 
%Since $f$ is continuous, it suffices to find a continuous $g\colon Z\rightarrow \NN^\NN$ with $g(Z)=B:=f^{-1}(A)$. 

Since analytic sets are preserved under Borel isomorphisms, $f^{-1}(A)$ is an analytic subset of $Y$ and thus of $\NN^\NN$. 
Let $B$ be a Borel subset of $\NN^\NN\times \NN^\NN$ with $f^{-1}(A)=p(B)$. 
Since Borel subsets of $\NN^\NN\times \NN^\NN$ are *analytic by Problem \ref{characterization of analytic sets} \ref{characterization of analytic sets - closure}, there is a closed subset $C$ of $(\NN^\NN)^3$ with $f^{-1}(A)=p(C)$. 
Since $(\NN^\NN)^3$ and  $\NN^\NN$ are homeomorphic, it is easy to see that there is a continuous surjection $h\colon \NN^\NN\rightarrow C$. Then $g=fph$ is as required. 
\end{proof} 






%%%%%%%%%%%%
\section{Borel measurable functions} 

A function $f\colon X\rightarrow Y$ between standard Borel spaces is called \emph{Borel measurable} if $f$-preimages of Borel sets are Borel. 
%A function $f\colon X\rightarrow Y$ is called \emph{Borel measurable} if for every Borel subset $C$ of $Y$, $f^{-1}(C)$ is a Borel subset of $X$. 

\begin{lemma} 
The following are equivalent for a function $f\colon X\rightarrow Y$: 
\begin{enumerate-(a)} 
\item 
$f$ is Borel measurable. 
\item 
The graph $G(f)$ of $f$ is a Borel subset of $X\times Y$. 
\end{enumerate-(a)} 
\end{lemma} 
\begin{proof} 
(a) $\Rightarrow$ (b): $f(x)=y$ $\Longleftrightarrow$ $x\in \bigcap_{k\in \NN} f^{-1}(N_{y{\upharpoonright}k})$ $\Longleftrightarrow$ $(x,y)\in \bigcap_{k\in\NN} \bigcup_{t\in \NN^k} f^{-1}(N_t) \times N_t$. 

(b) $\Rightarrow$ (a): 
It suffices to show that $f^{-1}(U)$ is Borel for any open subset of $Y$. 
By Proposition \ref{separation of analytic sets}, it is sufficient to show that both $f^{-1}(U)$ and its complement in $X$ are analytic. 
These claims follows from the equivalences $x\in f^{-1}(U) \Longleftrightarrow \exists y\in Y\ (x,y)\in G(f) \Longleftrightarrow \forall y\in Y\ ((x,y)\in G(f)\Rightarrow y\in U)$. 
\end{proof} 






\bibliographystyle{alpha}
\bibliography{references}
  
\end{document}
